---
title: "Final-explorations"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r tidyverse, include=FALSE}
library(tidyverse)
```

This document is a narrative of explorations of the Kaiser dataset, prior to the final paper.

Note: the data file `31118130.csv` is a symlink to the original file in the "../3 + 4" directory in this repository. 

```{r load raw_data}
raw_data <- read_csv("31118130.csv")
raw_data
```
Opening question: among survey respondents who respond "refuse to answer" to the question of whom they voted for for president, is there a detectable bias? Are liberals or conservatives, Biden or Trump or "other" voters, more likely to do this? Is there a way we can tell?

We can't know for sure, but the question is, can we make a data-supported argument, based on the data in this survey, that supports or rejects this idea?

First, let's get a breakdown of the responses to the question:

```{r}
table(raw_data$voted2)
# hat tip to https://stackoverflow.com/a/45386128/13603796
table(raw_data$voted2) %>% prop.table() %>% `*`(100) %>% round(2)
```
So 8.5% of respondents refused to answer the question. That's a significant amount, 1 in 12.

First, how do they lean? There are many possible variables we could look at; let's start with `ideology`:

```{r}
table(raw_data$ideology)
```

Let's break this down.

```{r}
ideology_by_refused_voted2 <- raw_data %>%
  select(ideology, voted2) %>%
  filter(voted2 == "Refused")
table(ideology_by_refused_voted2$ideology)
```

