---
title: "Homework 2 - Steve Linberg"
output:
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the tidyverse, as always:

```{r tidyverse, message = FALSE}
library(tidyverse)
```

Set the working directory to the directory containing this script, using the 
RStudio API (if available); if not, the WD should be set manually. The data
file is stored in the same directory as the script.

Probably overkill for this situation; the goal is to emulate Unix `$0` to make
the script portable rather than hardwiring in a working directory. (I'm a little
surprised that R doesn't make this easier...)

```{r setwd(), message = FALSE, results = "hide"}
if (rstudioapi::isAvailable()) {
  # https://stackoverflow.com/a/47045368/13603796
  setwd(dirname(rstudioapi::getSourceEditorContext()$path))
} else {
  # setwd(something_else)  # <- uncomment and specify
}
```

Load the tab-delimited file `covid-responses.tab.tsv`:

```{r "load_data", message = FALSE}
myData <- read_tsv("covid-responses.tab.tsv")
```

The resulting dataframe contains `r nrow(myData)` observations, and `r ncol(myData)` variables.